Graded Call-By-Push-Value

### 2020-07-08 popl21/popl21.tex
[ popl21 ] substitution

% https://tex.stackexchange.com/questions/502652/define-tcolorbox-in-math-mode
\usepackage{tcolorbox}
\newtcbox{\grayboxtext}{on line,colback=gray!20,colframe=gray!20,size=fbox,arc=0pt,boxrule=0pt}%,beforeafter skip=0pt,leftright skip=0pt}
\newcommand{\graybox}[1]{\grayboxtext{$#1$}}

@@

%\newcommand{}{\ensuremath{}}
\newcommand{\ie}{\emph{i.e.}}
\newcommand{\eg}{\emph{e.g.}}
\newcommand{\Eg}{\emph{E.g.}}
\newcommand{\loccit}{\emph{loc.\,cit.}}
\newcommand{\den}[2][]{\llbracket#2\rrbracket^{#1}}
\newcommand{\dent}[1]{\llparenthesis#1\rrparenthesis}
\newcommand{\To}{\ensuremath{\Rightarrow}}
%\newcommand{\todot}{\stackrel\cdot\to}
\newcommand{\todot}{\mathbin{\dot{\to}}}
\newcommand{\bN}{\ensuremath{\mathbb{N}}}
\newcommand{\dom}{\mathop{\mathrm{dom}}\nolimits}
\newcommand{\Pot}[1]{\mathcal{P}\,#1}
%\newcommand{\defiff}{:\iff}
\newcommand{\defiff}{\mathrel{{{:}{\Longleftrightarrow}}}}

@@

% Types
\newcommand{\tTy}{\mathsf{Ty}}
\newcommand{\Ty}[1]{\tTy^{#1}}
\newcommand{\PTy}{\Ty+}
\newcommand{\NTy}{\Ty-}
\newcommand{\Cxt}{\mathsf{Cxt}}
\newcommand{\RCxt}{\mathsf{RCxt}}
\newcommand{\cempty}{\emptyset}
\newcommand{\extr}[2]{#1.#2}
\newcommand{\ext}[3]{#1.#2{:}#3}
\newcommand{\qext}[5]{#1#2.#3#4{:}#5}
\newcommand{\gqext}[5]{\graybox{#1}#2.\graybox{#3}#4{:}#5}

@@

% Judgements
\newcommand{\ValTy}[3]{#1 \vdash #2 : #3}
\newcommand{\oCompTy}{\ValTy}
\newcommand{\CompTy}[4]{#1 \vdash #2 : #4 \mid #3}
\newcommand{\gpValTy}[4]{\graybox{(#1)}#2 \vdash #3 : #4}
\newcommand{\gqValTy}[4]{\graybox{#1}#2 \vdash #3 : #4}
\newcommand{\gpCompTy}[4]{\graybox{(#1)}#2 \vdash #3 : #4}
\newcommand{\gqCompTy}[4]{\graybox{#1}#2 \vdash #3 : #4}
\newcommand{\pValTy}[4]{(#1)#2 \vdash #3 : #4}
\newcommand{\qValTy}[4]{#1#2 \vdash #3 : #4}
\newcommand{\pCompTy}[4]{(#1)#2 \vdash #3 : #4}
\newcommand{\qCompTy}[4]{#1#2 \vdash #3 : #4}
%\newcommand{\CompTy}[4]{#1 \vdash #2 \mathrel{{:}\langle #3 \rangle} #4}
\newcommand{\ru}{\dfrac}
\newcommand{\nru}[3]{#1\;\dfrac{#2}{#3}}
\newcommand{\nrux}[4]{#1\;\dfrac{#2}{#3}\;#4}

\newcommand{\rulename}[1]{\ensuremath{\mbox{\textsc{#1}}}\xspace}
\newcommand{\rintro}[1]{\ensuremath{\mathord{#1}\mbox{-\rulename{intro}}}\xspace}
\newcommand{\relim}[1]{\ensuremath{\mathord{#1}\mbox{-\rulename{elim}}}\xspace}
\newcommand{\rvar}{\rulename{var}}
\newcommand{\rlet}{\rulename{let}}
\newcommand{\rsub}{\rulename{sub}}
\newcommand{\rweak}{\rulename{weak}}

@@

Negative types $N$ are interpreted
as functors $\denn N : (\Eff,\leq) \to \CC$ mapping
effect classifiers $e$ to objects $\Denn N e$ and effect subsumption $e
\leq e'$ to morphisms $\Denn N {e \leq e'} : \Denn N e \to \Denn N {e'}$.
\[
\begin{array}{lcl}
  \denp\_ & : & \PTy \to \CC \\
  \denp{\thunkty e N} & = & \Denn N e \\
  \denp{\sumt I P} & = & \coprod_{i:I} \denp{P_i} \\
  \denp{\tupt I P} & = & \prod_{i:I} \denp{P_i} \\
\end{array}
\qquad\qquad
\begin{array}{lcl}
  \Denn\_\_ & : & \NTy \to \Eff \to \CC \\
  \Denn{\diamond P} e & = & \T[e] \denp P \\
  \Denn{P \To N}   e & = & \denp P \To \Denn N e \\
  \Denn{\rect I N} e & = & \prod_{i:I} \Denn{N_i}e \\
  % \denn{P \To N}   e & = & \denp P \To (\denn N e) \\
  % \denn{\rect I N} e & = & \prod_{i:I} \left(\denn{N_i}e\right) \\
\end{array}
\]
The graded $\T$-algebra structure $\run{\denn N}$ is constructed by
induction on $N$:
\[
\begin{array}{lcl}
  \run{\denn N} & : & \Hom {\T[e_1](\Denn N {e_2})} {\Denn N {e_1 \bu e_2}} \\
  \run{\denn {\diamond P}} & = & \tjoin \\
  \run{\denn {P \To N}}   & = & \Lambda(\run{\denn N} \circ
                                \fmap[\T]\teval \circ \strengthr) \\
  \run{\denn {\rect I N}} & = & \langle \run{\denn{N_i}} \comp
                                \fmap[\T]{\pi_i} \rangle_{i:I} \\
\end{array}
\]
This construction is the same as \citet{levy:hosc06}, modulo grading.
The algebra laws ensue.

@@

As in Levy's algebra semantics \citeyearpar{levy:hosc06}, creation and
forcing of thunks is invisible in the model.  The bind operation
(\relim\diamond) utilizes the generalization of $\tjoin$ to $\trun$.
Subsumption is interpreted by functoriality in $\Eff$.
\[
\begin{array}{llcl}
\rintro\Box & \dent{\thunk t} & = & \dent t \\
\relim\Box  & \dent{\force v} & = & \dent v
\\[1ex]
\rintro\diamond & \dent{\ret v} & = & \treturn \comp \dent v \\
\relim\diamond  & \dent{\bind x u t} & = &
  \run{\denn N} \comp \fmap[\T]\teval \comp \strengthl \comp
  \langle \Lambda\dent t,\,\dent u \rangle
\\[1ex]
\rsub & \dent t & = & \Denn N {e \leq e'} \comp \dent t
\\
\end{array}
\]
These definitions are understood in the context of the given typing rules.

@@

\newcommand{\lolli}{\multimap}
\newcommand{\qfun}[3]{#1#2 \lolli #3}
\newcommand{\gqfun}[3]{\graybox{#1#2 \lolli #3}}
\renewcommand{\CompTy}[3]{#1 \vdash #2 : #3}
\newcommand{\nbox}[1]{}
\newcommand{\tzip}{\mathsf{zip}}
\newcommand{\textract}{\mathsf{extract}}
\newcommand{\extract}[1][]{\textract_{#1}}
\newcommand{\tdisplay}{\mathsf{display}}
\newcommand{\display}[1][]{\tdisplay_{#1}}
% \newcommand{\tduplicate}{\mathsf{display}}
% \newcommand{\duplicate}[1][]{\tdisplay_{#1}}
\newcommand{\tduplicate}{\mathsf{duplicate}}
\newcommand{\duplicate}[1][]{\tduplicate_{#1}}
\newcommand{\tcobind}{\mathsf{cobind}}
\newcommand{\tbind}{\mathsf{bind}}
\newcommand{\tget}{\mathsf{get}}
\newcommand{\tmodify}{\mathsf{modify}}
\newcommand{\thead}{\mathsf{head}}
\newcommand{\ttail}{\mathsf{tail}}
\newcommand{\suc}{(\_{+}1)}
\newcommand{\Float}{\mathsf{Float}}
\newcommand{\tiso}{\mathsf{iso}}
\newcommand{\isot}{\tiso^\otimes}

@@

      \CompTy {\qext \gamma \Gamma r y N} t {N'}

@@

      \CompTy {\gqext \gamma \Gamma r x P} t N

@@

      \CompTy {\gqext \gamma \Gamma r y N} t {N'}

@@

     {\gqValTy \delta \Gamma v {\sumt I P} \qquad
      \forall i{:}I,\ \CompTy {\gqext \gamma \Gamma r {x_i} {P_i}} {t_i} N

@@

@@

     {\gqValTy \delta \Gamma v {\tupt I P} \qquad
      \gqCompTy \delta {\extr \Gamma {\overline{\graybox{r}x_i{:}P_i}^{i:I}}} t N

@@

     {\CompTy {\gqext \gamma \Gamma r x P} t N

@@

Computations $\qCompTy \gamma \Gamma t N$ are interpreted as morphisms
$\dent t : \Hom {\Den\Gamma\gamma} {\denn N}$ and values
$\qValTy \gamma \Gamma v P$ as families
$\denv v r : \Hom {\Den\Gamma{r\gamma}} {\Denp P r}$ natural in $r$.
Naturality here means that
$\Denp P {r \leq s} \comp \denv v r = \denv v s \comp \left(\Den \Gamma {(\_
    \cdot \gamma)}\right)_{r \leq s}$.
Again, because of $\rweak$, we interpret typing derivations rather than terms.
The interpretation is now rather straightforward, but we spell it out for reference.
\[
\begin{array}{llcl}
\rvar^+
  & \denv {\ValTy{\qext 0 \Gamma 1 x P} x P} r
  & = & \Gr \comp (\tdrop[\den\Gamma] \otimes \tid[\Denp P r])
\\
\rintro\Box
  & \denv {\qValTy \gamma \Gamma {\thunk t} {\Box N}} r
  & = & \fmapt[{\D[r]}]{\dent t} \comp \expose{\den \Gamma}
\\
\rintro\GS
  & \denv{\qValTy \gamma \Gamma {\inj i v} {\sumt I P}} r
  & = & \iota_i \comp \denv v r
\\
\rintro\otimes
  & \denv{\pValTy {\sumt I \gamma} \Gamma {\tup v} {\tupt I P}} r
  & = & \tupt {i:I} \denv {v_i}r \comp \duplicate
\\
\end{array}
\]
\[
\begin{array}{llcl}
\rvar^-
  & \dent {\oCompTy{\qext 0 \Gamma 1 y N} y N}
  & = & \Gr \comp (\tdrop[\den\Gamma] \otimes \extract[\denn N])
\\
\rlet^{-}
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qlet r y u t} {N'}}
  & = & \dent t
  \comp (\tid[\Den \Gamma \gamma] \otimes
           (\fmapt[{\D[r]}] \dent u \comp \expose[\den \Gamma]))
\\ &&& {~}
  \comp \duplicate
\\
\rlet^{+}
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qlet r x v t} {N}}
  & = & \dent t
  \comp (\tid[\Den \Gamma \gamma] \otimes
           (\expose[\denp P] \comp \denv v r))
  \comp \duplicate
\\
\relim\Box
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qforced r v y t} {N'}}
  & = & \dent t
  \comp (\tid[\Den \Gamma \gamma] \otimes \denv v r)
  \comp \duplicate
\\
\relim\GS
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\caser v \{r x_i.\, t_i\}_{i:I}} {N}}
  & = & [\dent {t_i}]_{i:I} \comp \tdist
  \comp (\tid[\Den \Gamma \gamma] \otimes \denv v r)
  \comp \duplicate
\\
\relim\otimes
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qsplits r v {\bar x} t} {N}}
  & = & \dent t \comp \isot
  \comp (\tid[\Den \Gamma \gamma] \otimes \denv v r)
  \comp \duplicate
\\
\rintro\diamond
  & \dent {\pCompTy {r\gamma} \Gamma {\ret v} {\compty r P}}
  & = & \treturn \comp \denv v r
\\
\relim\diamond
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qbind r x u t} {N}}
  & = & \run[\den N] \comp \fmapt[\T]{\dent t} \comp \strengthl
\\ &&& {~}
  \comp (\tid[\Den \Gamma \gamma] \otimes \dent u)
  \comp \duplicate
\\
\rintro\lolli
  & \dent {\qCompTy \gamma \Gamma {\lam x t} {\qfun r P N}}
  & = & \Lambda \dent t
\\
\relim\lolli
  & \dent{\pCompTy {\gamma + r\delta} \Gamma {\app t v} N}
  & = & \teval \comp (\dent t \otimes \denv v r) \comp \duplicate
\\
\rintro\Pi
  & \dent {\qCompTy \gamma \Gamma {\recrd I t} {\rect I N}}
  & = & \langle \dent {t_i} \rangle_{i:I}
\\
\relim\Pi
  & \dent {\qCompTy \gamma \Gamma {\proj i t} {N_i}}
  & = & \pi_i \comp \dent t
\\
\rweak
  & \dent {\qCompTy \gamma \Gamma t N}
  & = & \dent t \comp \Den \Gamma {\gamma \leq \gamma'}
\\
\end{array}
\]

>>>>

% https://tex.stackexchange.com/questions/502652/define-tcolorbox-in-math-mode
\usepackage{tcolorbox}
\newtcbox{\grayboxtext}{on line,colback=gray!50,colframe=gray!50,size=fbox,arc=0pt,boxrule=0pt}%,beforeafter skip=0pt,leftright skip=0pt}
\newcommand{\graybox}[1]{\grayboxtext{$#1$}}

@@

%\newcommand{}{\ensuremath{}}
\newcommand{\ie}{\emph{i.e.}}
\newcommand{\eg}{\emph{e.g.}}
\newcommand{\Eg}{\emph{E.g.}}
\newcommand{\loccit}{\emph{loc.\,cit.}}
\newcommand{\cf}{cf.\ }
\newcommand{\den}[2][]{\llbracket#2\rrbracket^{#1}}
\newcommand{\dent}[1]{\llparenthesis#1\rrparenthesis}
\newcommand{\To}{\ensuremath{\Rightarrow}}
%\newcommand{\todot}{\stackrel\cdot\to}
\newcommand{\todot}{\mathbin{\dot{\to}}}
\newcommand{\bN}{\ensuremath{\mathbb{N}}}
\newcommand{\dom}{\mathop{\mathrm{dom}}\nolimits}
\newcommand{\Pot}[1]{\mathcal{P}\,#1}
%\newcommand{\defiff}{:\iff}
\newcommand{\defiff}{\mathrel{{{:}{\Longleftrightarrow}}}}
\newcommand{\subst}[3]{#3[#1/#2]}
\newcommand{\qsubst}[4]{#4[#1/#2#3]}

@@

% Types
\newcommand{\tTy}{\mathsf{Ty}}
\newcommand{\Ty}[1]{\tTy^{#1}}
\newcommand{\PTy}{\Ty+}
\newcommand{\NTy}{\Ty-}
\newcommand{\Cxt}{\mathsf{Cxt}}
\newcommand{\RCxt}{\mathsf{RCxt}}
\newcommand{\cempty}{\emptyset}
\newcommand{\sempty}{\emptyset}
\newcommand{\sext}[3]{#1,#3/#2}
\newcommand{\extr}[2]{#1.#2}
\newcommand{\ext}[3]{#1.#2{:}#3}
\newcommand{\qext}[5]{#1#2.#3#4{:}#5}
\newcommand{\gqext}[5]{\graybox{#1}#2.\graybox{#3}#4{:}#5}

@@

% Judgements
\newcommand{\SubstTy}[4]{#1#2 \vdash #3 : #4}
\newcommand{\pSubstTy}[4]{(#1)#2 \vdash (#3) : (#4)}
\newcommand{\ValTy}[3]{#1 \vdash #2 : #3}
\newcommand{\oCompTy}{\ValTy}
\newcommand{\CompTy}[4]{#1 \vdash #2 : #4 \mid #3}
\newcommand{\gpValTy}[4]{\graybox{(#1)}#2 \vdash #3 : #4}
\newcommand{\gqValTy}[4]{\graybox{#1}#2 \vdash #3 : #4}
\newcommand{\gpCompTy}[4]{\graybox{(#1)}#2 \vdash #3 : #4}
\newcommand{\gqCompTy}[4]{\graybox{#1}#2 \vdash #3 : #4}
\newcommand{\pValTy}[4]{(#1)#2 \vdash #3 : #4}
\newcommand{\qValTy}[4]{#1#2 \vdash #3 : #4}
\newcommand{\pCompTy}[4]{(#1)#2 \vdash #3 : #4}
\newcommand{\qCompTy}[4]{#1#2 \vdash #3 : #4}
%\newcommand{\CompTy}[4]{#1 \vdash #2 \mathrel{{:}\langle #3 \rangle} #4}
\newcommand{\ru}{\dfrac}
\newcommand{\nru}[3]{#1\;\dfrac{#2}{#3}}
\newcommand{\nrux}[4]{#1\;\dfrac{#2}{#3}\;#4}

\newcommand{\rulename}[1]{\ensuremath{\mbox{\textsc{#1}}}\xspace}
\newcommand{\rbeta}[1]{\ensuremath{\beta\mbox{-}\mathord{#1}}\xspace}
\newcommand{\reta}[1]{\ensuremath{\eta\mbox{-}\mathord{#1}}\xspace}
\newcommand{\rpi}[1]{\ensuremath{\pi\mbox{-}\mathord{#1}}\xspace}
\newcommand{\rintro}[1]{\ensuremath{\mathord{#1}\mbox{-\rulename{intro}}}\xspace}
\newcommand{\relim}[1]{\ensuremath{\mathord{#1}\mbox{-\rulename{elim}}}\xspace}
\newcommand{\rvar}{\rulename{var}}
\newcommand{\rlet}{\rulename{let}}
\newcommand{\rsub}{\rulename{sub}}
\newcommand{\rweak}{\rulename{weak}}

@@

Negative types $N$ are interpreted
as functors $\denn N : (\Eff,\leq) \to \CC$ mapping
effect classifiers $e$ to objects $\Denn N e$ and effect subsumption $e
\leq e'$ to morphisms $\Denn N {e \leq e'} : \Denn N e \to \Denn N {e'}$.
\[
\begin{array}{lcl}
  \denp\_ & : & \PTy \to \CC \\
  \denp{\thunkty e N} & = & \Denn N e \\
  \denp{\sumt I P} & = & \coprod_{i:I} \denp{P_i} \\
  \denp{\tupt I P} & = & \prod_{i:I} \denp{P_i} \\
\end{array}
\qquad\qquad
\begin{array}{lcl}
  \Denn\_\_ & : & \NTy \to \Eff \to \CC \\
  \Denn{\diamond P} e & = & \T[e] \denp P \\
  \Denn{P \To N}   e & = & \denp P \To \Denn N e \\
  \Denn{\rect I N} e & = & \prod_{i:I} \Denn{N_i}e \\
  % \denn{P \To N}   e & = & \denp P \To (\denn N e) \\
  % \denn{\rect I N} e & = & \prod_{i:I} \left(\denn{N_i}e\right) \\
\end{array}
\]
The graded $\T$-algebra structure $\run[\denn N]$ is constructed by
induction on $N$:
\[
\begin{array}{lcl}
  \run[\denn N] & : & \Hom {\T[e_1](\Denn N {e_2})} {\Denn N {e_1 \bu e_2}} \\
  \run[\denn {\diamond P}] & = & \tjoin \\
  \run[\denn {P \To N}]   & = & \Lambda(\run[\denn N] \circ
                                \fmap[\T]\teval \circ \strengthr) \\
  \run[\denn {\rect I N}] & = & \langle \run[\denn{N_i}] \comp
                                \fmap[\T]{\pi_i} \rangle_{i:I} \\
\end{array}
\]
This construction is the same as \citet{levy:hosc06}, modulo grading.
The algebra laws ensue.

@@

As in Levy's algebra semantics \citeyearpar{levy:hosc06}, creation and
forcing of thunks is invisible in the model.  The bind operation
(\relim\diamond) utilizes the generalization of $\tjoin$ to $\trun$.
Subsumption is interpreted by functoriality in $\Eff$.
\[
\begin{array}{llcl}
\rintro\Box & \dent{\thunk t} & = & \dent t \\
\relim\Box  & \dent{\force v} & = & \dent v
\\[1ex]
\rintro\diamond & \dent{\ret v} & = & \treturn \comp \dent v \\
\relim\diamond  & \dent{\bind x u t} & = &
  \run[\denn N] \comp \fmap[\T]\teval \comp \strengthl \comp
  \langle \Lambda\dent t,\,\dent u \rangle
\\[1ex]
\rsub & \dent t & = & \Denn N {e \leq e'} \comp \dent t
\\
\end{array}
\]
These definitions are understood in the context of the given typing rules.

@@

\newcommand{\lolli}{\multimap}
\newcommand{\qfun}[3]{#1#2 \lolli #3}
\newcommand{\gqfun}[3]{\graybox{#1#2 \lolli #3}}
%\renewcommand{\CompTy}[3]{#1 \vdash #2 : #3}
\newcommand{\nbox}[1]{}
\newcommand{\tzip}{\mathsf{zip}}
\newcommand{\textract}{\mathsf{extract}}
\newcommand{\extract}[1][]{\textract_{#1}}
\newcommand{\tdisplay}{\mathsf{display}}
\newcommand{\display}[1][]{\tdisplay_{#1}}
% \newcommand{\tduplicate}{\mathsf{display}}
% \newcommand{\duplicate}[1][]{\tdisplay_{#1}}
\newcommand{\tduplicate}{\mathsf{duplicate}}
\newcommand{\duplicate}[1][]{\tduplicate_{#1}}
\newcommand{\tcobind}{\mathsf{cobind}}
\newcommand{\tbind}{\mathsf{bind}}
\newcommand{\tget}{\mathsf{get}}
\newcommand{\tmodify}{\mathsf{modify}}
\newcommand{\thead}{\mathsf{head}}
\newcommand{\ttail}{\mathsf{tail}}
\newcommand{\suc}{(\_{+}1)}
\newcommand{\Float}{\mathsf{Float}}
\newcommand{\tiso}{\mathsf{iso}}
\newcommand{\isot}{\tiso^\otimes}

@@

      \oCompTy {\qext \gamma \Gamma r y N} t {N'}

@@

      \oCompTy {\gqext \gamma \Gamma r x P} t N

@@

      \oCompTy {\gqext \gamma \Gamma r y N} t {N'}

@@

     {\gqValTy \delta \Gamma v {\sumt I P} \quad
      \forall i{:}I,\ \oCompTy {\gqext \gamma \Gamma r {x_i} {P_i}} {t_i} N

@@

\quad

@@

     {\gqValTy \delta \Gamma v {\tupt I P} \quad
      \gqCompTy \gamma {\extr \Gamma {\overline{\graybox{r}x_i{:}P_i}^{i:I}}} t N

@@

     {\oCompTy {\gqext \gamma \Gamma r x P} t N

@@

Computations $\qCompTy \gamma \Gamma t N$ are interpreted as morphisms
$\dent t : \Hom {\Den\Gamma\gamma} {\denn N}$ and values
$\qValTy \gamma \Gamma v P$ as families
$\denv v r : \Hom {\Den\Gamma{r\gamma}} {\Denp P r}$ natural in $r$.
Naturality here means that
$\Denp P {r \leq s} \comp \denv v r = \denv v s \comp \left(\Den \Gamma {(\_
    \cdot \gamma)}\right)_{r \leq s}$.
Again, because of $\rweak$, we interpret typing derivations rather than terms.
The interpretation is now rather straightforward, but we spell it out for reference.
\[
\begin{array}{llcl}
\rvar^+
  & \denv {\ValTy{\qext 0 \Gamma 1 x P} x P} r
  & = & \Gr \comp (\tdrop[\den\Gamma] \otimes \tid[\Denp P r])
\\
\rintro\Box
  & \denv {\qValTy \gamma \Gamma {\thunk t} {\Box N}} r
  & = & \fmapt[{\D[r]}]{\dent t} \comp \expose[\den \Gamma]
\\
\rintro\GS
  & \denv{\qValTy \gamma \Gamma {\inj i v} {\sumt I P}} r
  & = & \iota_i \comp \denv v r
\\
\rintro\otimes
  & \denv{\pValTy {\sumt I \gamma} \Gamma {\tup v} {\tupt I P}} r
  & = & \tupt {i:I} \denv {v_i}r \comp \duplicate
\\
\end{array}
\]
\[
\begin{array}{llcl}
\rvar^-
  & \dent {\oCompTy{\qext 0 \Gamma 1 y N} y N}
  & = & \Gr \comp (\tdrop[\den\Gamma] \otimes \extract[\denn N])
\\
\rlet^{-}
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qlet r y u t} {N'}}
  & = & \dent t
  \comp (\tid[\Den \Gamma \gamma] \otimes
           (\fmapt[{\D[r]}] \dent u \comp \expose[\den \Gamma]))
\\ &&& {~}
  \comp \duplicate
\\
\rlet^{+}
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qlet r x v t} {N}}
  & = & \dent t
  \comp (\tid[\Den \Gamma \gamma] \otimes \denv v r)
  \comp \duplicate
\\
\relim\Box
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qforced r v y t} {N'}}
  & = & \dent t
  \comp (\tid[\Den \Gamma \gamma] \otimes \denv v r)
  \comp \duplicate
\\
\relim\GS
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\caser v \{r x_i.\, t_i\}_{i:I}} {N}}
  & = & [\dent {t_i}]_{i:I} \comp \tdist
  \comp (\tid[\Den \Gamma \gamma] \otimes \denv v r)
  \comp \duplicate
\\
\relim\otimes
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qsplits r v {\bar x} t} {N}}
  & = & \dent t \comp \isot
  \comp (\tid[\Den \Gamma \gamma] \otimes \denv v r)
  \comp \duplicate
\\
\rintro\diamond
  & \dent {\pCompTy {r\gamma} \Gamma {\ret v} {\compty r P}}
  & = & \treturn \comp \denv v r
\\
\relim\diamond
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qbind r x u t} {N}}
  & = & \run[\den N] \comp \fmapt[\T]{\dent t} \comp \strengthl
\\ &&& {~}
  \comp (\tid[\Den \Gamma \gamma] \otimes \dent u)
  \comp \duplicate
\\
\rintro\lolli
  & \dent {\qCompTy \gamma \Gamma {\lam x t} {\qfun r P N}}
  & = & \Lambda \dent t
\\
\relim\lolli
  & \dent{\pCompTy {\gamma + r\delta} \Gamma {\app t v} N}
  & = & \teval \comp (\dent t \otimes \denv v r) \comp \duplicate
\\
\rintro\Pi
  & \dent {\qCompTy \gamma \Gamma {\recrd I t} {\rect I N}}
  & = & \langle \dent {t_i} \rangle_{i:I}
\\
\relim\Pi
  & \dent {\qCompTy \gamma \Gamma {\proj i t} {N_i}}
  & = & \pi_i \comp \dent t
\\
\rweak
  & \dent {\qCompTy \gamma \Gamma t N}
  & = & \dent t \comp \Den \Gamma {\gamma \leq \gamma'}
\\
\end{array}
\]

It is remarkable that coeffect-graded CBPV works \emph{without any
  distributive law} \cite{orchard:icfp16}
for the monad $\T$ and the graded comonad $\D$.
%
This was a crucial design criterion, leading to resource qualifiers in
the monadic type $\compty r P$.
%
A distributive law would be required if we allowed scaling of
computations, not just of values.

\subsection{Coeffect-graded CBPV: equational theory and operational semantics}

The equational theory of pure CBPV comprises $\beta$ and $\eta$ laws,
and \emph{sequencing laws} that syntactically express that negative
types are monad algebras \citep[Fig.~11]{levi:hosc06}.  The sequencing
laws contain a generalization of the associativity law for monads and
allow to permute a bind under a $\lambda$ or record construction.  In
the presence of $\eta$ for functions and records, they are
interderivable with the following permutation laws:
\[
\begin{array}{llcl}
\rpi\diamond
  & \bind {x_2} {(\bind {x_1} {t_1} {t_2})} {t_3}
  & = & \bind {x_1} {t_1} {(\bind {x_2} {t_2} {t_3})}
\\
\rpi\lolli
  & \app {(\bind x u t)} v
  & = & \bind x u {\app t v}
\\
\rpi\Pi
  & \proj i {(\bind x  u t)}
  & = & \bind x u {\proj i t}
\end{array}
\]
Note that permutations for let-bindings like
$\app{(\qlet r x v t)}w = \qlet r x v {\app t w}$ are instances of the
$\beta$-law $\qlet r x v t = \qsubst v r x t$.
%
Herein, \fbox{$\qsubst v r x t$} shall the denote the
(capture-avoiding) replacement $\subst v x t$ of a positive variable
$x$ annotated by $r$ by value $v$ in term $t$.
%
Permutations for value-eliminations like
$\app {(\qsplits r v {\bar x} t)} w = \qsplits r v {\bar x} {\app t
  w}$
are derivable (using the $\beta$-laws) from the $\eta$-laws, like
$\qsubst{v}{r}{z}{t'} = \qsplits r {v} {\bar x} {\qsubst {\ptup {\bar
      x}} r z {t'}}$ for $\otimes$.

Our extension of CBPV features negative variables $y$ standing for
computations $u$, thus, we require a notion of computation
substitution $\qsubst u r t y$ that should have the following semantics.
\begin{align*}
&
  \ru{\qCompTy \delta \Gamma u N \qquad
      \oCompTy {\qext \gamma \Gamma r y N} t {N'}
    }{\pCompTy {\gamma + r\delta} \Gamma {\qsubst u r t y} {N'}}
\\ &
  \dent{\qsubst u r t y} = \dent t
    \comp (\tid[\den\Gamma \gamma]
             \otimes (\fmapt[{\D[r]}] \dent u \comp \expose[\den\Gamma]))
    \comp \duplicate[\den\Gamma]
\end{align*}

Further, we add $\beta$-laws for the negative
let-binding and $\relim\Box$, since $\tforce$ now has let-style elimination:
\[
\begin{array}{llcl}
\rbeta{\rlet^{-}}
  & \qlet r y u t
  & = & \qsubst u r y t
\\
\rbeta\Box
  & \qforced r {\thunk u} y t
  & = & \qsubst u r y t
\\
\end{array}
\]
These laws are immediately sound, since left and right hand sides of
these equations have the same denotation.
%
Read as reduction rules, these laws are added to the operational
semantics of CBPV, replacing $\beta$-law $\force{(\thunk u)} = u$.

Substitution for coeffect-graded type system has been worked
out in detail by \citet{atkeyWood:types19}.  It straightforwardly extends to
coeffect-graded CBPV.
%
A substitution $\sigma$ is a finite map from variable to terms,
mapping positive variables $x$ to values $v$ and negative variables
$y$ to computations $t$.  Substitution typing
\fbox{$\SubstTy \Psi \Gamma \sigma \Delta$} is equipped with a matrix
$\Psi \in \R^{\dom\Gamma \times \dom\Delta}$ recording the usage
vector $\Psi(z) \in \R^{\dom\Gamma}$ for each variable
$z \in \dom\Delta = \dom \sigma$ in the domain of the substitution.
\[
  \ru{
    }{\SubstTy {0} \Gamma \sempty \cempty}
\qquad
  \ru{\SubstTy \Psi \Gamma \sigma \Delta \qquad
      \qValTy \gamma \Gamma v P
    }{\pSubstTy {\Psi,\gamma} \Gamma {\sext \sigma x v} {\ext \Delta x P}}
\qquad
  \ru{\SubstTy \Psi \Gamma \sigma \Delta \qquad
      \qCompTy \gamma \Gamma t N
    }{\pSubstTy {\Psi,\gamma} \Gamma {\sext \sigma y t} {\ext \Delta y N}}
\]
Capture-avoiding parallel substitution into values \fbox{$v \sigma$} and
computations \fbox{$t \sigma$} is defined as usual by recursion on the term.

The matrix $\Psi : \R^{\dom \Gamma \times \dom \Delta}$ acts as a
linear map $\R^{\dom \Delta} \to \R^{\dom \Gamma}$ and allows us to
state the substitution theorem.
\begin{theorem}[Substitution preserves typing]
  Let $\SubstTy \Psi \Gamma \sigma \Delta$.
  \begin{enumerate}
  \item If $\qValTy \delta \Delta v P$ then $\pValTy {\Psi\delta} \Gamma {v\sigma} P$.
  \item If $\qCompTy \delta \Delta t N$ then
    $\pCompTy {\Psi\delta} \Gamma {t\sigma} N$.
  \end{enumerate}
\end{theorem}
\begin{proof}
  Straightforward adaptation of \citet{atkeyWood:types19}.
\end{proof}
\begin{corollary}[Subject reduction]
  $\beta$-reduction of coeffect-graded CBPV preserves types.
\end{corollary}
In the pure (effect-free) version, every computation at value type
returns a value.
\begin{conjecture}[Canonicity of pure coeffect-graded CBPV]
  If $\vdash t : \diamond P$ then $t =_{\beta} \ret v$ for some $\vdash v : P$.
\end{conjecture}
This normalization result can be proved with standard techniques such
as reducibility candidates or normalization by evaluation (\cf
\citet{abelSattler:ppdp19}).

The denotation of a substitution $\SubstTy \Psi \Gamma \sigma \Delta$
is a natural transformation
$\denv \sigma \delta : \Hom {\Den \Gamma {\Psi\delta}} {\Den \Delta
  \delta}$
such that
$
\Den \Delta {\delta \leq \delta'}
\comp
\denv \sigma \delta
=
\denv \sigma {\delta'}
\comp
\Den \Gamma {\Psi\delta \leq \Psi\delta'}
$.
It is defined by recursion as follows:
\begin{align*}
& \ru{
    }{\SubstTy {0} \Gamma \sempty \cempty}
& \dent \sempty
& : \Hom {\Den\Gamma 0} \I
& & = \tdrop[\den\Gamma]
\\[2ex]
& \ru{\SubstTy \Psi \Gamma \sigma \Delta \qquad
      \qValTy \gamma \Gamma v P
    }{\pSubstTy {\Psi,\gamma} \Gamma {\sext \sigma x v} {\ext \Delta x P}}
& \dent{\sext \sigma x v}
& : \Hom  {\Den\Gamma{\Psi\delta,r\gamma}} {\Den \Delta \delta \otimes \Denp P r}
& & = \denv \sigma \delta \otimes \denv v r
\\[2ex]
& \ru{\SubstTy \Psi \Gamma \sigma \Delta \qquad
      \qCompTy \gamma \Gamma t N
    }{\pSubstTy {\Psi,\gamma} \Gamma {\sext \sigma y t} {\ext \Delta y N}}
\end{align*}

\section{Fully graded CBPV}
\label{sec:full-cbpv}

\newcommand{\peCompTy}[5]{(#1)#2 \vdash #3 : #5 \mid #4}
\newcommand{\qeCompTy}[5]{#1#2 \vdash #3 : #5 \mid #4}
% \newcommand{\eext}[4]{#1.#2{:}#4#3}
% \newcommand{\qeext}[6]{#1#2.#3#4{:}#6#5}
\newcommand{\eext}[4]{#1\mathbin.#2:#4|#3}
\newcommand{\qeext}[6]{#1#2\mathbin.#3#4:#6|#5}
